Weighted Model Counting with Conditional Weights for Bayesian Networks

### 2021-06-04 doc/long_talk/talk.tex
Finishing touches to the UAI submission

\begin{frame}[fragile]{Weighted Model Counting (WMC)}
  \begin{columns}
    \begin{column}{0.5\textwidth}
      \begin{itemize}
      \item Generalises propositional model counting ($\#\SAT{}$)
      \item Applications:
        \begin{itemize}
        \item graphical models
        \item probabilistic programming
        \item neural-symbolic artificial intelligence
        \end{itemize}
      \item Main types of algorithms:
        \begin{itemize}
        \item using knowledge compilation
        \item using a \SAT{} solver
        \item manipulating pseudo-Boolean functions
        \end{itemize}
      \end{itemize}
    \end{column}
    \begin{column}{0.5\textwidth}
      \begin{example}
      $w(x) = 0.3$, $w(\neg x) = 0.7$, $w(y) = 0.2$, $w(\neg y) = 0.8$
      \vspace{1cm}

@@

\begin{frame}{The Limitations and Capabilities of Classical WMC}
  \begin{alertblock}{Observation}
    Classical WMC is only able to evaluate \structure{factorable} measures
    (c.f., a collection of mutually independent random variables).
  \end{alertblock}
  \begin{theorem}[Informal Version]
    It is always possible to add more variables to turn a non-factorable measure
    into a factorable measure.
  \end{theorem}
  However, that is not necessarily a good idea!
\end{frame}

\begin{frame}{Conditional Probability Tables (CPTs) as Functions}
  \begin{itemize}
  \item \alert{One} variable for every random variable with \alert{two} values.
  \item \alert{$k$} variables for every random variable with \alert{$k > 2$}
    values.
  \item Define indicator functions of the form \structure{$[x]\colon 2^{\{x\}}
      \to \{ 0, 1 \}$}.
    \begin{itemize}
    \item \structure{$[x](\emptyset) = 0$};
    \item \structure{$[x](\{ x \}) = 1$}.
    \end{itemize}
  \item Define \structure{$+$}, \structure{$\cdot$}, and scalar multiplication pointwise.
  \item Then a CPT can be represented as a function.
  \end{itemize}
  \begin{columns}[t]
    \begin{column}{0.5\textwidth}
      \centering
      \begin{tabular}[t]{ccc}
        \toprule
        $a$ & $b$ & $\Pr(A = a \mid B = b)$ \\
        \midrule
        1 & 1 & 0.6 \\
        1 & 0 & 0.4 \\
        0 & 1 & 0.1 \\
        0 & 0 & 0.9 \\
        \bottomrule
      \end{tabular}
    \end{column}
    \begin{column}{0.5\textwidth}
      \begin{align*}
        \mathsf{CPT_A} &= 0.6[\lambda_{A=1}] \cdot [\lambda_{B=1}] \\
                       &+ 0.4[\lambda_{A=1}] \cdot \overline{[\lambda_{B=1}]} \\
                       &+ 0.1\overline{[\lambda_{A=1}]} \cdot [\lambda_{B=1}] \\
                       &+ 0.9\overline{[\lambda_{A=1}]} \cdot \overline{[\lambda_{B=1}]},
      \end{align*}
    \end{column}
  \end{columns}
\end{frame}

@@

\begin{frame}{Summary and Future Work}
  \begin{itemize}
  \item Classical WMC can represent any probability distribution by adding more
    variables.
  \item But this is not the right approach for WMC algorithms that support
    working directly with functions.
  \item Specifically with ADDMC, avoiding redundant variables resulted in
    \alert{$127$} times faster inference.
  \item Could this idea be successfully applied to other applications of WMC or,
    perhaps, other WMC algorithms?
  \end{itemize}
  \begin{itemize}
  \item Potential improvements to the encoding:
    \begin{itemize}
    \item Apply ideas from other WMC encodings for Bayesian networks (e.g.,
      prime implicants, $\log$ encoding).
    \item Develop encoding tricks that apply to functions but not to conjunctive
      normal form.
    \item More on this in our SAT~2021 paper \emph{\structure{Weighted Model
          Counting Without Parameter Variables}}.
    \end{itemize}
  \end{itemize}
\end{frame}

>>>>

\begin{frame}[fragile]{Weighted Model Counting (WMC)}
  \begin{columns}
    \begin{column}{0.5\textwidth}
      \begin{itemize}
      \item Generalises propositional model counting ($\#\SAT{}$)
      \item Applications:
        \begin{itemize}
        \item graphical models
        \item probabilistic programming
        \item neural-symbolic artificial intelligence
        \end{itemize}
%      \item Main types of algorithms:
%        \begin{itemize}
%        \item using knowledge compilation
%        \item using a \SAT{} solver
%        \item manipulating pseudo-Boolean functions
%        \end{itemize}
      \end{itemize}
    \end{column}
    \begin{column}{0.5\textwidth}
      \begin{example}
      $w(x) = 0.3$, $w(\neg x) = 0.7$, $w(y) = 0.2$, $w(\neg y) = 0.8$
      \vspace{1cm}

@@

\begin{frame}{The Limitations and Capabilities of WMC}
  \begin{alertblock}{Observation}
    Classical WMC is only able to evaluate \structure{factorable} measures
    (c.f., a collection of mutually independent random variables).
  \end{alertblock}
  \begin{theorem}[Informal Version]
    It is always possible to add more variables to turn a non-factorable measure
    into a factorable measure.
  \end{theorem}
  However, that is not necessarily a good idea!
\end{frame}

\begin{frame}{Encoding Bayesian Networks}
  \begin{itemize}
%  \item \alert{One} variable for every random variable with \alert{two} values.
%  \item \alert{$k$} variables for every random variable with \alert{$k > 2$}
%    values.
  \item Define indicator functions of the form \structure{$[x]\colon 2^{\{x\}}
      \to \{ 0, 1 \}$}.
    \begin{itemize}
    \item \structure{$[x](\emptyset) = 0$};
    \item \structure{$[x](\{ x \}) = 1$}.
    \end{itemize}
  \item Define \structure{$+$}, \structure{$\cdot$}, and scalar multiplication pointwise.
  \item Then a conditional probability table (CPT) can be represented as a function.
  \end{itemize}
  \begin{columns}[t]
    \begin{column}{0.5\textwidth}
      \centering
      \begin{tabular}[t]{ccc}
        \toprule
        $a$ & $b$ & $\Pr(A = a \mid B = b)$ \\
        \midrule
        1 & 1 & 0.6 \\
        1 & 0 & 0.4 \\
        0 & 1 & 0.1 \\
        0 & 0 & 0.9 \\
        \bottomrule
      \end{tabular}
    \end{column}
    \begin{column}{0.5\textwidth}
      \begin{align*}
        \mathsf{CPT_A} &= 0.6[\lambda_{A=1}] \cdot [\lambda_{B=1}] \\
                       &+ 0.4[\lambda_{A=1}] \cdot \overline{[\lambda_{B=1}]} \\
                       &+ 0.1\overline{[\lambda_{A=1}]} \cdot [\lambda_{B=1}] \\
                       &+ 0.9\overline{[\lambda_{A=1}]} \cdot \overline{[\lambda_{B=1}]},
      \end{align*}
    \end{column}
  \end{columns}
\end{frame}

@@

\begin{frame}{Summary and Future Work}
  \begin{itemize}
  \item (Classical) WMC can represent any probability distribution by adding
    more variables.
  \item But this is not the right approach for WMC algorithms that support
    working directly with functions.
  \item Specifically with ADDMC, avoiding redundant variables resulted in
    \alert{$127$} times faster inference.
%  \item Could this idea be successfully applied to other applications of WMC or,
%    perhaps, other WMC algorithms?
  \end{itemize}
  \begin{itemize}
  \item Potential improvements to the encoding:
    \begin{itemize}
    \item Apply ideas from other WMC encodings for Bayesian networks (e.g.,
      prime implicants, $\log$ encoding).
    \item Develop encoding tricks that apply to functions but not to conjunctive
      normal form.
    \item More on this in our SAT~2021 paper \emph{\structure{Weighted Model
          Counting Without Parameter Variables}}.
    \end{itemize}
  \end{itemize}
\end{frame}
